Distributional Regression Forests for Probabilistic Precipitation Forecasting in Complex Terrain

### 2018-03-21 pkg/disttree/inst/draft/draft.Rnw
small corrections
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2068 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 23275f833564eea728d6714a822b1b4fc2d89988

To embed the distributional approach from GAMLSS into regression trees and random
forests, we proceed in three steps. (1)~To fix notation, we briefly review fitting
distributions using standard maximum likelihood in in Section~\ref{sec:distfit}.
(2)~A recursive partitioning strategy based on the corresponding scores (or
gradients) is introduced in Section~\ref{sec:disttree}, leading to distributional
trees. (3)~Ensembles of distributional trees fitted to randomized subsamples are
employed to establish distributional forests in Section~\ref{sec:distforest}.

The general distributional notation is exemplified in all three steps for the
zero-censored Gaussian distribution. The latter is employed in the empirical
in the empirical case study in Section~\ref{sec:precipitation} to model
power-transformed daily precipitation amounts.

@@

To capture both location and scale of the probabilistic precipitation forecasts
while accounting for a point mass at zero precipitation (i.e., dry days without
rain), zero-censored Gaussian distribution with with mean/location parameter
$\mu$ and standard deviation/scale parameter $\sigma$ is employed. Therefore,
the corresponding log-likelihood function with parameter vector 
$\bm{\theta} = (\mu, \sigma)$ is
$$
\ell(\mu, \sigma; Y) = 
\begin{cases}
    \log\left\{\frac{1}{\sigma} \cdot \phi\left(\frac{Y - \mu}{\sigma}\right) \right\}, & \text{if } Y > 0\\[0.2cm]
    \log\left\{\Phi\left(\frac{-\mu}{\sigma}\right)\right\}, & \text{if } Y = 0
\end{cases}
$$
where $\phi$ and $\Phi$ are the probability density function and the 
distribution function of the standard normal distribution $\mathcal{N}(0,1)$. 
Other distributions $\mathcal{D}$ and corresponding log-likelihoods
$\ell(\mu, \sigma; Y)$ could be set up in the same way, e.g., for
censored shifted gamma distributions \citep{Scheuerer+Hamill:2015} or
zero-censored logistic distributions \citep{Gebetsberger+Messner+Mayr:2017}.

@@

In summary, distributional trees are fitted recursively via:
\begin{enumerate}
\item Estimate $\bm{\hat \theta}$ via maximum likelihood for the observations
  in the current subsample.
\item Test for associations or instabilities of the scores $s(\bm{\hat \theta}, y_i)$ 
  and $Z_{l,i}$ for each partitioning variable~$Z_l$ ($l = 1, \dots, m$).
\item Split the sample along the partitioning variable $Z_{l,i}^*$ with the
  strongest association or instability. Choose the breakpoint with the highest
  improvement in the log-likelihood or the highest discrepancy .
\item Repeat steps 1--3 recursively in the subsamples until these become too
  small or there is no significant association/instability (or some other
  stopping criterion is reached).
\end{enumerate}
%
Different inference techniques can be used for assessing the association between
scores and covariates in step~3. In the following we use the general class of
permutation tests introduced by \cite{Hothorn+Hornik+VanDeWiel:2006} which is
also the basis of conditional inference trees \citep[CTree,][]{Hothorn+Hornik+Zeileis:2006}.
Alternatively, one could use asymptotic M-fluctuation tests for parameter
instability \citep{Zeileis+Hornik:2007} as in model-based recursive partitioning
\citep[MOB,][]{Zeileis+Hothorn+Hornik:2008}. More details are provided in
Appendix~\ref{app:tree}.

@@

Many weather forecasting models leverage the strengths of modern
numerical weather prediction (NWP) systems \citep[see][]{Bauer+Thorpe+Brunet:2015}
and produce high-quality forecasts adapted to specific sites/locations
through distributional regression models like EMOS
(ensemble model output statistics, \citealp{Gneiting+Raftery+Westveld:2005}).
In the case of precipiation forecasting, EMOS explains past precipitation
sums by the NWP output ``total precipitation'' (\emph{tp}), using its
ensemble mean as the predictor variable for the location parameter and
the ensemble standard deviation for the scale parameter.

@@

In the following, it is illustrated how distributional regression forests
can solve this problem of automatically selecting relevant variables,
interactions, and potentially nonlinear effects. For fitting the forest
only the response distribution and the list of potential predictor variables
need to be specified (along with a few algorithmic details) and then
the model fit is determined by the forest in a data-driven way. Here, we
employ a zero-censored Gaussian distribution and 80~predictor variables
computed from ensemble means and spreads of various NWP outputs.
The predictive performance of the forest is compared to three other
zero-censored Gaussian models: (a) a standard basic EMOS, (b) a GAMLSS
with prespecified effects and interactions based on meteorological
knowedge/experience, and (c) a boosted GAMLSS with automatic
selection of smooth additive terms based on all 80~predictor variables.

@@

To assess predictive performance, a full cross-validation is carried out
rather than relying on just the one test set for the years 2009--2012.
To do so, 10 times 7-fold cross-validation is carried out where each
splits the available 28~years into 7~subsets of 4~randomly selected years.
The models learned on 6~folds (= 24~years) and evaluated on the 7-th fold
(= 4~years) using the average CRPS across all observations.
The resulting 10~CRPS skill scores are displayed by boxplots in
Figure~\ref{fig:axams-crps} using EMOS as the reference model (horizontal
line at a CRPSS of 0). Clearly, both GAMLSS models and the distributional forest 
perform better than the EMOS model. While the two GAMLSS lead to an improvement 
of around 4~percent, the distributional forest has a slightly higher improvement
of around 5.5~percent.

@@

As an alternative to QQ~plots based on randomized quantile residuals,
PIT (probability integral transform) histograms 
\cite{Gneiting+Balabdaoui+Raftery:2007} are a commonly used to 
assess how well-calibrated distributional fits are.
Figure~\ref{fig:pit} shows out-of-sample PIT histograms for the same
probabilistic forecasts as Figure~\ref{fig:qq}, confirming that
despite some deviations the overall calibration of all models is acceptable.

>>>>

To embed the distributional approach from GAMLSS into regression trees and random
forests, we proceed in three steps. (1)~To fix notation, we briefly review fitting
distributions using standard maximum likelihood in Section~\ref{sec:distfit}.
(2)~A recursive partitioning strategy based on the corresponding scores (or
gradients) is introduced in Section~\ref{sec:disttree}, leading to distributional
trees. (3)~Ensembles of distributional trees fitted to randomized subsamples are
employed to establish distributional forests in Section~\ref{sec:distforest}.

The general distributional notation is exemplified in all three steps for the
zero-censored Gaussian distribution. The latter is employed in the empirical 
case study in Section~\ref{sec:precipitation} to model power-transformed daily 
precipitation amounts.

@@

To capture both location and scale of the probabilistic precipitation forecasts
while accounting for a point mass at zero precipitation (i.e., dry days without
rain), zero-censored Gaussian distribution with mean/location parameter
$\mu$ and standard deviation/scale parameter $\sigma$ is employed. Therefore,
the corresponding log-likelihood function with parameter vector 
$\bm{\theta} = (\mu, \sigma)$ is
$$
\ell(\mu, \sigma; Y) = 
\begin{cases}
    \log\left\{\frac{1}{\sigma} \cdot \phi\left(\frac{Y - \mu}{\sigma}\right) \right\}, & \text{if } Y > 0\\[0.2cm]
    \log\left\{\Phi\left(\frac{-\mu}{\sigma}\right)\right\}, & \text{if } Y = 0
\end{cases}
$$
where $\phi$ and $\Phi$ are the probability density function and the 
distribution function of the standard normal distribution $\mathcal{N}(0,1)$. 
Other distributions $\mathcal{D}$ and corresponding log-likelihoods
$\ell(\mu, \sigma; Y)$ could be set up in the same way, e.g., for
censored shifted gamma distributions \citep{Scheuerer+Hamill:2015} or
zero-censored logistic distributions \citep{Gebetsberger+Messner+Mayr:2017}.

@@

In summary, distributional trees are fitted recursively via:
\begin{enumerate}
\item Estimate $\bm{\hat \theta}$ via maximum likelihood for the observations
  in the current subsample.
\item Test for associations or instabilities of the scores $s(\bm{\hat \theta}, y_i)$ 
  and $Z_{l,i}$ for each partitioning variable~$Z_l$ ($l = 1, \dots, m$).
\item Split the sample along the partitioning variable $Z_l^*$ with the
  strongest association or instability. Choose the breakpoint with the highest
  improvement in the log-likelihood or the highest discrepancy .
\item Repeat steps 1--3 recursively in the subsamples until these become too
  small or there is no significant association/instability (or some other
  stopping criterion is reached).
\end{enumerate}
%
Different inference techniques can be used for assessing the association between
scores and covariates in step~3. In the following we use the general class of
permutation tests introduced by \cite{Hothorn+Hornik+VanDeWiel:2006} which is
also the basis of conditional inference trees \citep[CTree,][]{Hothorn+Hornik+Zeileis:2006}.
Alternatively, one could use asymptotic M-fluctuation tests for parameter
instability \citep{Zeileis+Hornik:2007} as in model-based recursive partitioning
\citep[MOB,][]{Zeileis+Hothorn+Hornik:2008}. More details are provided in
Appendix~\ref{app:tree}.

@@

Many weather forecasting models leverage the strengths of modern
numerical weather prediction (NWP) systems \citep[see][]{Bauer+Thorpe+Brunet:2015}
and produce high-quality forecasts adapted to specific sites/locations
through distributional regression models like EMOS
(ensemble model output statistics, \citealp{Gneiting+Raftery+Westveld:2005}).
In the case of precipitation forecasting, EMOS explains past precipitation
sums by the NWP output ``total precipitation'' (\emph{tp}), using its
ensemble mean as the predictor variable for the location parameter and
the ensemble standard deviation for the scale parameter.

@@

In the following, it is illustrated how distributional regression forests
can solve this problem of automatically selecting relevant variables,
interactions, and potentially nonlinear effects. For fitting the forest
only the response distribution and the list of potential predictor variables
need to be specified (along with a few algorithmic details) and then
the model fit is determined by the forest in a data-driven way. Here, we
employ a zero-censored Gaussian distribution and 80~predictor variables
computed from ensemble means and spreads of various NWP outputs.
The predictive performance of the forest is compared to three other
zero-censored Gaussian models: (a) a standard basic EMOS, (b) a GAMLSS
with prespecified effects and interactions based on meteorological
knowledge/experience, and (c) a boosted GAMLSS with automatic
selection of smooth additive terms based on all 80~predictor variables.

@@

To assess predictive performance, a full cross-validation is carried out
rather than relying on just the one test set for the years 2009--2012.
To do so, 10 times 7-fold cross-validation is carried out where each
splits the available 28~years into 7~subsets of 4~randomly selected years.
The models are learned on 6~folds (= 24~years) and evaluated on the 7-th fold
(= 4~years) using the average CRPS across all observations.
The resulting 10~CRPS skill scores are displayed by boxplots in
Figure~\ref{fig:axams-crps} using EMOS as the reference model (horizontal
line at a CRPSS of 0). Clearly, both GAMLSS models and the distributional forest 
perform better than the EMOS model. While the two GAMLSS lead to an improvement 
of around 4~percent, the distributional forest has a slightly higher improvement
of around 5.5~percent.

@@

As an alternative to QQ~plots based on randomized quantile residuals,
PIT (probability integral transform) histograms 
\cite{Gneiting+Balabdaoui+Raftery:2007} are a commonly used tool to 
assess how well-calibrated distributional fits are.
Figure~\ref{fig:pit} shows out-of-sample PIT histograms for the same
probabilistic forecasts as Figure~\ref{fig:axams-qq}, confirming that
despite some deviations the overall calibration of all models is acceptable.
